Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-03-31 response-to-reviews.Rmd

line481 'is maximized' or is maximal? Please correct.

**We have replaced “maximal” with “greatest”.**

line485-487: It seems that VPD and its effect on gs is ignored here. Please clarify? The well-known stomatal closure at high VPD alone would mean higher Tair would decrease gs in the upper canopy (not 'canopy leaves' as stated) compared to the lower canopy. See Grossiord et al. (2020) Tansley Review. VPD only gets mentioned in the section on photosynthesis (l. 496). Please do so earlier.

**We agree, and have modified the text in this section to emphasize the role of VPD in driving stomatal closure in sun leaves (new text italicized): “However, ~~because~~ cannot meet the demands caused by the high radiation and $g_b$ in sun leaves ***with fully open stomata, because of height-related constraints on water transport (e.g., Yoder et al. 1994, Koch et al. 2004, Sillett et al 2010) and/or increased leaf-air VPD caused by leaf warming in sunlit canopy locations [@buckleyStomatalConductanceOptimized2014]***. These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table 2), which drives the lower intracellular CO~2~ and $\delta^{13}C$ discussed in section 3.3 (Table 1).  Stomatal depression reduces transpirational cooling (Fig. 3), thus amplifying the warming of sun leaves by high radiation. ~~In sun leaves, Tleaf thus further increases due to the lack of transpirational cooling.~~” We have also added the phrase ***“(driven by increased VPD)”***after “sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$”, and clarified that “high $T_{air}$ should decrease $g_s$ of ***upper canopy***leaves”.**

@@

line650: 'In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover'. Having studied large overstory trees and young understory saplings, I have a very, very difficult time with this statement especially portraying it as categorically true even if consistent with the 4 cited studies. Very few researchers have studies the same species in this context, and in this review manuscript the authors cannot scientifically lay the effect on canopy position and microclimate rather than age or other confounding factors. I must challenge their assertion and ask a higher level of evidence to make these statements, which should be qualified.

**As the point about resilience was not critical, we have removed that part of the statement. The sentence now reads, "Drought also causes greater mortality in larger trees [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019]."**

line946: '… how … temperature sensitivities of metabolism and woody growth vary across these vertical gradients'. I’m not sure why the authors are speculating on woody growth in this context, surely it would be extremely difficult to show that woody growth varied vertically in an individual? And that across-individual differences compound comparisons of individuals in different vertical strata? Shouldn’t these things be pointed out to readers?

**This has been reworded to read, "However, uncertainty remains about how temperature sensitivity of foliar gas exchange varies across these vertical gradients. Similarly, much remains unknown about how crown exposure influences the temperature sensitivity of woody stem growth. While most available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig. 4)."**

### 2022-03-31 vertical-thermal-review.Rmd
moved 2.1 to notes

### 2.1. Biophysical drivers of $T_{leaf}$

Fundamentally, $T_{leaf}$ is determined by the energy balance and can be estimated based on biophysical principles: $T_{leaf}-T_{air}$ depends on energy input from net radiation ($R_n$, including shortwave and longwave) minus heat lost to the environment [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019].
High $R_n$ loads can elevate $T_{leaf}$ above $T_{air}$ (Fig. `r fig_leaf_T`a). 
Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance, which is greater in smaller leaves (Fig. `r fig_leaf_T`d) and under higher wind speeds (Fig. `r fig_leaf_T`b). 
Latent heat flux ($\lambda E$) through transpiration cools the leaf, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD:$g_b$ increases with wind speed, and $g_s$ generally declines as VPD increases due to stomatal closure [@darwinIXObservationsStomata1898; @mottStomatalResponsesHumidity1991].
Therefore, $T_{leaf}-T_{air}$ decreases in magnitude with wind speed [Fig. `r fig_leaf_T`b, @daudetWindSpeedLeaf1999], increases with RH (Fig. `r fig_leaf_T`c), increases with leaf size (Fig. `r fig_leaf_T`d), and $T_{leaf}$ decreases absolutely with $g_s$ (Fig. `r fig_leaf_T`e).

@@

Under hot and dry conditions, leaves face a trade-off between $T_{leaf}$ regulation and water conservation [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
With adequate water, high $\lambda E$ can dissipate excess heat, particularly for smaller leaves (higher $g_b$) [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021a; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
However, when transpiration exceeds water supply, stomata close to conserve water, increasing $T_{leaf}$ [Fig. `r fig_leaf_T`e, @fauset_differences_2018].
Therefore, at high solar radiation loads, latent cooling can maintain $T_{leaf}$ closer to $T_{air}$, but if stomatal opening is limited, solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021a].

### 2.2 Vertical gradients in leaf temperature

The bbiophysical principles outlined above determine how $T_{leaf}$ is above $T_{air}$ across vertical gradients in closed-canopy forests.
High radiation in the upper canopy (Fig. `r fig_NEON_vertical`) implies that upper canopy leaves often have higher $T_{leaf}-T_{air}$, and thus are often warmer than understory leaves (Fig. `r fig_leaf_T`).
However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce leaf-air temperature differences [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see sections 3, 4.1) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig. `r fig_leaf_T`b-e). 
In contrast, in the lower canopy, greater RH and lower wind speeds may limit $\lambda E$ (through reduced VPD and $g_b$), increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
Leaves in these normally-shaded locations experience rapid light and temperature surges during sunflecks (>10$^\circ$C)  [@waySunflecksTreesForests2012].

@@

>>>>

@@

### 2.1 Vertical gradients in leaf temperature

Biophysical principles (`r Notes_S1`) determine how elevated $T_{leaf}$ is above $T_{air}$ across vertical gradients in closed-canopy forests.
High radiation in the upper canopy (Fig. `r fig_NEON_vertical`) implies that upper canopy leaves often have higher $T_{leaf}-T_{air}$, and thus are often warmer than understory leaves (Fig. `r fig_leaf_T`).
However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce leaf-air temperature differences [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see sections 3, 4.1) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig. `r fig_leaf_T`b-e). 
In contrast, in the lower canopy, greater RH and lower wind speeds may limit $\lambda E$ (through reduced VPD and $g_b$), increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
Leaves in these normally-shaded locations experience rapid light and temperature surges during sunflecks (>10$^\circ$C)  [@waySunflecksTreesForests2012].

@@

`r Notes_S1`

### 2022-03-31 vertical-thermal-review_SI.Rmd
moved 2.1 to notes

# SI
## appendices

@@

>>>>

# SI
##Notes

Notes_S1<- "Notes S1"

## appendices

@@

# `r Notes_S1`: Biophysical drivers of $T_{leaf}$

Fundamentally, $T_{leaf}$ is determined by the energy balance and can be estimated based on biophysical principles: $T_{leaf}-T_{air}$ depends on energy input from net radiation ($R_n$, including shortwave and longwave) minus heat lost to the environment [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019].
High $R_n$ loads can elevate $T_{leaf}$ above $T_{air}$ (Fig. 3a). 
Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance, which is greater in smaller leaves (Fig. 3d) and under higher wind speeds (Fig. 3b). 
Latent heat flux ($\lambda E$) through transpiration cools the leaf, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD:$g_b$ increases with wind speed, and $g_s$ generally declines as VPD increases due to stomatal closure [@darwinIXObservationsStomata1898; @mottStomatalResponsesHumidity1991].
Therefore, $T_{leaf}-T_{air}$ decreases in magnitude with wind speed [Fig. 3b, @daudetWindSpeedLeaf1999], increases with RH (Fig. `r fig_leaf_T`c), increases with leaf size (Fig. `r fig_leaf_T`d), and $T_{leaf}$ decreases absolutely with $g_s$ (Fig. 3e).

Under hot and dry conditions, leaves face a trade-off between $T_{leaf}$ regulation and water conservation [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
With adequate water, high $\lambda E$ can dissipate excess heat, particularly for smaller leaves (higher $g_b$) [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021a; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
However, when transpiration exceeds water supply, stomata close to conserve water, increasing $T_{leaf}$ [Fig. 3e, @fauset_differences_2018].
Therefore, at high solar radiation loads, latent cooling can maintain $T_{leaf}$ closer to $T_{air}$, but if stomatal opening is limited, solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021a].
